---
title: "lucas_analysis"
output: html_document
---

```{r loading packages}
suppressWarnings(suppressPackageStartupMessages(library(tidyverse)))
suppressWarnings(suppressPackageStartupMessages(library(survival)))

mode_stat <- function(v) names(which.max(table(v)))
```

```{r loading data}
actg_c <- readRDS('../data/actg_clean.rds')
```

```{r KM Curves}
surv_actg <- Surv(actg_c$time_event, actg_c$event)

survfit_actg <- survfit(surv_actg ~ tx, data = actg_c)
survfit_na_actg <- survfit(surv_actg ~ tx, data = actg_c, type = 'fleming-harrington')

survfit_actg |> 
  plot(main = "Disease Free Survival for Soc and Invidir",
       col = 1:2, lwd = 2,
       conf.int = TRUE,
       xlab = 'Days',
       ylab = 'Survival Prob.',
       ylim = c(0.7, 1))
legend('bottomleft', c('SoC', 'SoC + Invidir'), col = 1:2, lwd = 2)

survfit_actg |> 
  plot(main = 'KM vs. AE Curve',
       col = 1:2, lwd = 2,
       xlab = 'Days',
       ylab = 'Survival Prob',
       ylim = c(0.7, 1))
lines(survfit_na_actg, col = 3:4)
legend('bottomleft', c('KM Soc', 'KM SoC + Invidir', 'AE SoC', 'AE SoC + Invidir'), col = c(1:4), lwd = 1)
```

We see the AE and KM curves to be identical

```{r testing hypothesis}
survdiff(surv_actg ~ tx, data = actg_c)
```

From the Mantzel-Haeznel Test, we see there to be a difference in the two treatments. We should investigate this with cox model for proportional hazards.

```{r small model}
small_cox_actg <- coxph(surv_actg ~ tx, data = actg_c)
summary(small_cox_actg)
```

```{r large model}
large_cox_actg <- coxph(surv_actg ~ tx + sex + hemophil + ivdrug + karnof + priorzdv + age + race + base_cd4, data = actg_c)
summary(large_cox_actg)
```

Lets see if proportionality holds for this large model, and then investigate residuals if needed to tighten it up before we start dropping variables, or adding to the large one. 

```{r checking proportionality}
plot(survfit_na_actg,
     col = 1:2, lwd = 2,
     fun = 'cloglog',
     main = 'Complimentary Log-Log Survival Curves')
legend('topleft', c('SoC', 'SoC + Invidir'), col = 1:2, lwd = 2)

plot(survfit_na_actg, col = 1:2, lwd = 2, ylim = c(0.7, 1), main = 'Observed vs. Expected Survival (Small Model)',
     xlab = 'Days',
     ylab = 'Survival Prob.')
lines(survfit(small_cox_actg, data.frame(tx = c(0, 1)), data = actg_c),
      col = 1:2, lwd = 2, lty = 2)
legend('bottomleft', c('SoC', 'SoC + Invidir'), col = 1:2, lwd = 2)

large_model_df <- data.frame(tx = c(0, 1),
                             sex = c('male', 'male'),
                             ivdrug = c('never', 'never'),
                             hemophil = c(0, 0),
                             karnof = c(mode_stat(actg_c$karnof[actg_c$tx == 0]), mode_stat(actg_c$karnof[actg_c$tx == 1])),
                             priorzdv = c(median(actg_c$priorzdv[actg_c$tx == 0]), median(actg_c$priorzdv[actg_c$tx == 1])),
                             age = c(mean(actg_c$age[actg_c$tx == 0]), mean(actg_c$age[actg_c$tx == 1])),
                             race = c(mode_stat(actg_c$race[actg_c$tx == 0]), mode_stat(actg_c$race[actg_c$tx == 1])),
                             base_cd4 = c(median(actg_c$base_cd4[actg_c$tx == 0]), median(actg_c$base_cd4[actg_c$tx == 1])))

plot(survfit_na_actg, col = 1:2, lwd = 2, ylim = c(0.7, 1), main = 'Observed vs. Expected Survival (Large Model)',
     xlab = 'Days',
     ylab = 'Survival Prob.')
lines(survfit(large_cox_actg, large_model_df, data = actg_c),
      col = 1:2, lwd = 2, lty = 2)
legend('bottomleft', c('SoC', 'SoC + Invidir'), col = 1:2, lwd = 2)
```

We see that proportionality seems to hold in our cloglog plot. Looking at the observed vs. expected survival curves, we see that a small model fits the observed survival much better than our large model does, pointing in the direction that a small model will be better. Drop1 will likely not be very effective, as there will be a lot to change, ie combinations of dropping, so we will instead try add1, and wremove variables that seem obviously unhelpful, and then investigate residuals to see which variables are causing problems and then remove them from the model, as well as which variables can be modified to keep our assumptions strong (linearity, etc.).

```{r add1}
add1(small_cox_actg, scope = ~ tx + sex + hemophil + ivdrug + karnof + priorzdv + age + race + base_cd4, test = "Chisq")
```

We see that keeping the karnofsky score, base_cd4, and age are helpful to the model, ivdrug use and priorzdv are borderline, so we will keep those as well and see if there are any transformations we can work with. We also have clinical reasons to believe priorzdv may be helpful in prediction, as it's a crucial part in the standard of care

```{r}
mid_cox_actg <- coxph(surv_actg ~ tx + ivdrug + karnof + priorzdv + age + base_cd4, data = actg_c)
summary(mid_cox_actg)
```

```{r schoenfeld residuals for proportionality}
mid_model_zph <- 
  cox.zph(mid_cox_actg)

mid_model_zph

plot(mid_model_zph[1],
     main = 'Plot of Schoenfeld Residuals',
     xlab = 'Time',
     ylab = 'Treatment')

plot(mid_model_zph[2],
     main = 'Plot of Schoenfeld Residuals',
     xlab = 'Time',
     ylab = 'IVDrug')

plot(mid_model_zph[3],
     main = 'Plot of Schoenfeld Residuals',
     xlab = 'Time',
     ylab = 'Karnofsky Score')

plot(mid_model_zph[4],
     main = 'Plot of Schoenfeld Residuals',
     xlab = 'Time',
     ylab = 'Prior ZDV Use (Months)')

plot(mid_model_zph[5],
     main = 'Plot of Schoenfeld Residuals',
     xlab = 'Time',
     ylab = 'Age')

plot(mid_model_zph[6],
     main = 'Plot of Schoenfeld Residuals',
     xlab = 'Time',
     ylab = 'Base_cd4 (mm^3)')
```

None of the variables bring up super apparent proportionality issues graphically, but based on p-values, base_cd4 is borderline.

```{r martingale residuals for transformations}
mres_priorzdv <- residuals(coxph(surv_actg ~ tx + ivdrug + karnof + age + base_cd4, data = actg_c))
mres_age <- residuals(coxph(surv_actg ~ tx + ivdrug + karnof + priorzdv + base_cd4, data = actg_c))
mres_basecd4 <- residuals(coxph(surv_actg ~ tx + ivdrug + karnof + priorzdv + age, data = actg_c))

plot(actg_c$priorzdv, mres_priorzdv,
     xlab = 'Months of Prior ZDV',
     ylab = 'Martingale Residuals',
     main = 'Martingale Residuals vs. ZDV Usage')
lines(lowess(actg_c$priorzdv, mres_priorzdv))

plot(actg_c$age, mres_age,
     xlab = 'Age',
     ylab = 'Martingale Residuals',
     main = 'Martingale Residuals vs. Age')
lines(lowess(actg_c$age, mres_age))

plot(actg_c$base_cd4, mres_basecd4,
     xlab = 'Base cd4',
     ylab = 'Martingale Residuals',
     main = 'Martingale Residuals vs. Base cd4')
lines(lowess(actg_c$base_cd4, mres_basecd4))
```

No linear transformations necessary it looks like from the martingale residuals at least

```{r cox-snell residuals for goodness of fit}
mart_actg <- residuals(mid_cox_actg, type = 'martingale')
cs_actg <- actg_c$event - mart_actg

surv_cs <- survfit(Surv(cs_actg, actg_c$event) ~ 1, type = 'fleming-harrington')
plot(surv_cs, fun = 'cumhaz',
     main = 'Cumulative Hazard of Cox-Snell Residuals')
abline(0, 1, col = 'red')
```

We don't see a lack of fit from this procedure. There don't seem to be any obvious ways of changing the large model to get a consistently better fit. Lastly, lets investigate interaction terms between variables.

It seems logical that age and karnofsky score would have an interaction. A 60 year olf with a score of 100 is likely less resiliant to disease than a 25 year old with a score of 100.

```{r karnofsky score and age}
summary(mid_cox_actg)
summary(coxph(surv_actg ~ tx + ivdrug + priorzdv + base_cd4 + (age * karnof), data = actg_c))

mres_mid <- residuals(mid_cox_actg, type = 'martingale')

plot((actg_c$age * as.numeric(actg_c$karnof)), mres_mid,
     xlab = 'Age * Karnofsky Score',
     ylab = 'Martingale Residuals',
     main = 'Martingale Residuals vs. Age * Karnofsky Score')
lines(lowess((actg_c$age * as.numeric(actg_c$karnof)), mres_mid))
```

That was a miss, its likely that the intuition is fine, but actually age and karnofsky score are highly correlated. Since this was a short study it's likely that overall health is more important that age. In the long run, age may cause complications with having a combined event of death and AIDS, but not here. Lets check covariance.

```{r cov of karnof and age and result}
boxplot(age ~ karnof, data = actg_c,
        main = 'Boxplots of Age by Karnofsky Score',
        xlab = 'Karnofsky Score',
        ylab = 'Age')
```

Those with the lower score are slightly older, but not by much to assume that there is great co-linearity between them, we should be fine keeping both variables.

Next, priorzdv may be a proxy for how long someone has had HIV, which could be an important measure in AIDS progression, but different to base_cd4. Lets see if they have high co-linearity.

```{r}
plot(actg_c$priorzdv, actg_c$base_cd4,
     xlim = c(0, 200),
     xlab = 'Prior ZDV',
     ylab = 'Base cd4',
     main = 'Plot of Base cd4 against Prior ZDV')
```

Again, there is no real pattern here between someone's baseline cd4 level and their prior months of ZDV usage, so there is no reason to believe they are having a strong effect on each other.

We can try now to see if an interaction term here is helpful, its reasonable that the effect of someones prior zdv usage on their survival depends on their baseline cd4 as someone with a higher cd4 should have a better time with the disease, but longer zdv implies longer time with HIV, which has detrimental effects on health outside of just cd4.

```{r}
summary(mid_cox_actg)
summary(coxph(surv_actg ~ tx + ivdrug + (priorzdv * base_cd4) + age + karnof, data = actg_c))

plot((actg_c$priorzdv * actg_c$base_cd4), mres_mid,
     xlab = 'ZDV * Base cd4',
     ylab = 'Martingale Residuals',
     main = 'Martingale Residuals vs. ZDV Usage * Base cd4')
lines(lowess((actg_c$priorzdv * actg_c$base_cd4), mres_mid))
```

Again, no dice, an interaction term helps us not here.

```{r checking survival of mid model against tx model}
mid_df <- data.frame(tx = c(0, 1),
                     ivdrug = c('never', 'never'),
                     karnof = c(mode_stat(actg_c$karnof[actg_c$tx == 0]), mode_stat(actg_c$karnof[actg_c$tx == 1])),
                     priorzdv = c(median(actg_c$priorzdv[actg_c$tx == 0]), median(actg_c$priorzdev[actg_c$tx == 1])),
                     age = c(mean(actg_c$age[actg_c$tx == 0]), mean(actg_c$age[actg_c$tx == 1])),
                     base_cd4 = c(median(actg_c$base_cd4[actg_c$tx == 0]), median(actg_c$base_cd4[actg_c$tx == 1])))

plot(survfit_na_actg, col = 1:2, lwd = 2, ylim = c(0.7, 1), main = 'Observed vs. Expected Survival (Mid Model)',
     xlab = 'Days',
     ylab = 'Survival Prob.')
lines(survfit(mid_cox_actg, mid_df, data = actg_c),
      col = 1:2, lwd = 2, lty = 2)
legend('bottomleft', c('SoC', 'SoC + Invidir'), col = 1:2, lwd = 2)

plot(survfit_na_actg, col = 1:2, lwd = 2, ylim = c(0.7, 1), main = 'Observed vs. Expected Survival (Large Model)',
     xlab = 'Days',
     ylab = 'Survival Prob.')
lines(survfit(large_cox_actg, large_model_df, data = actg_c),
      col = 1:2, lwd = 2, lty = 2)
legend('bottomleft', c('SoC', 'SoC + Invidir'), col = 1:2, lwd = 2)
```

Checking this mid-model against the larger model, we see that there is no discernible difference. We can't find any linearity issues in our variables, and there are no obvious combined effects to create to help with the model.

Summary so far: MAYBE base_cd4 violates the proportional hazards assumption, as we saw from its schoenfeld residuals, but overall model fit is pretty good, and none of the continuous covariates need a transformation it seems. Look at stratification because of the base_cd4, and look for outliars with dfbeta and deviance residuals.

Also: May want to remove those few values of individuals who use iv drugs currently, they are comprimising their health in a dramatic way, and they add an unnecessary variable to the model